---
title: "Covid19_Report"
author: "Siva"
date: '2022-07-22'
output: html_document
---

## COVID 19 Data Analysis Report

### Source of the data

COVID 19 Data is loaded from the  Johns Hopkins github site <https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series>. The dataset is read in from the following urls
<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv>, 
<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv> ,
<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv> , 
<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv> , 
<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>.

### Description of data

These are the datasets from the repository for the 2019 Novel Coronavirus Visual Dashboard operated by the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE).
These datasets records worldwide and US wide COVID-19 cases, deaths alongwith population,country, provice etc on a daily basis.


```{r setup, echo=TRUE}
# loading libraries
library(tidyverse)
library(dplyr)
library(lubridate)
#library(readr)
#library(stringr)
```


```{r load_url, echo=TRUE}
# loading url

url1 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
url2 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"

url3 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"
url4 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"
url5 <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
  
```

```{r read_data, echo=TRUE}
# read in data
US_cases <- read_csv(url1)
global_cases <- read_csv(url2)
US_deaths <- read_csv(url3)
global_deaths <- read_csv(url4)
lookup <- read_csv(url5)
```

```{r clean_data, echo=TRUE}
# Data cleaning for global data
global_cases <- global_cases %>% pivot_longer(cols=-c("Province/State","Country/Region","Lat" , "Long"),names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))   

global_deaths <- global_deaths %>% pivot_longer(cols=-c("Province/State","Country/Region","Lat" , "Long"),names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long)) 

# Joining global cases and deaths in one table
global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region="Country/Region", Province_State=`Province/State`) %>% mutate(date=mdy(date))

global <- global %>% filter(cases>0)

```

```{r  echo=TRUE, results='asis'}
library(knitr)
kable(global[1:5, ], caption="Global cases and deaths")
```

```{r clean_data_US, echo=TRUE}

US_cases <- US_cases %>% pivot_longer(cols=-(UID:Combined_Key),names_to = "date",values_to = "cases") %>% select(Admin2:cases) %>% mutate(date=mdy(date)) %>% select(-c(Lat,Long_))

US_deaths <- US_deaths %>% pivot_longer(cols=-(UID:Combined_Key),names_to = "date",values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date=mdy(date)) %>% select(-c(Lat,Long_))

US <- US_cases %>%  full_join(US_deaths)

```


```{r  add_population , echo = TRUE }
# Add population to the global table
global <- global %>%  unite("Combined_key",c(Province_State,Country_Region),sep=', ',na.rm=TRUE,remove=FALSE)

global <- global %>% left_join(lookup,by=c("Province_State", "Country_Region")) %>% select(-c(UID,FIPS)) %>% select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)

```


### Problem Statement 1

Trying to find out the Covid 19 cases and deaths in India from January 2020 till July 2022. 

```{r  plot , echo = TRUE }

india_plot <- global %>% filter(Country_Region=="India")

ggplot(india_plot,aes(x=date,y=cases)) + geom_line(aes(color="cases")) + geom_point(aes(color="cases")) + geom_line(aes(y=deaths,color="deaths")) + scale_y_log10() + scale_x_date(date_labels="%b %Y", breaks = "1 month")+ theme(legend.position="bottom", axis.text.x = element_text(angle =90)) +labs(title="Covid19 in India", y=NULL)

```


### Problem Statement 2

Trying to find out number of cases and deaths every month in India.

```{r  dmplot , echo = TRUE }
da <- india_plot %>% select(date,cases,deaths) %>% mutate(month=month(date))
da <- india_plot %>% select(date,cases,deaths) %>% mutate(month=month(date)) %>% select(month,cases,deaths) %>% group_by(month)

da1 <- da %>% group_by(month) %>% mutate(cases=sum(cases), deaths=sum(deaths)) %>% distinct()
pt <- da1 %>% gather(key,value,cases,deaths) %>% ggplot(aes(x=month,y=value,color=key)) +geom_point()+scale_y_log10()+ggtitle("Number of cases and deaths by month in India")+xlab("Month")

pt

```


```{r  US_by_states , echo = TRUE , eval=FALSE}

US <- US %>% left_join(lookup,by=c("Province_State", "Country_Region")) %>% select(-c(UID,FIPS)) %>% select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)

US_by_state <- US %>%  group_by(Province_State,Country_Region,date) %>% summarise(cases=sum(cases),deaths=sum(deaths),Population=sum(Population)) %>% mutate(deaths_per_million=deaths *1000000/Population) %>% select(Province_State,Country_Region,date,cases,deaths,deaths_per_million,Population) %>% ungroup()

```

### Source of bias

Bias may be in data collection. As a huge population some small cities and its data could have been ignored or not recorded. Confirmation bias can occur as heavy manual labour is required in recording the data.

### Further Investigations

Is big cities receive good healthcare rather than villages?
Is cases higher in urban or rural areas?
Is deaths higher in rural or urban areas?


### Conclusions

As we see the plot cases and deaths aren't equal. There is a huge gap in these two variables. 

```{r session_info, echo=TRUE}
sessionInfo()
```